---
title: "BIGpopA Tutorial"
subtitle: "Pedigree cleaning, validation, and parentage assignment from SNP genotype data"
author: "Breeding Insight - Josue Chinchilla-Vargas"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    css: BIcolors_template.css
    includes:
      in_header: header.html
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    number_sections: true
    code_folding: show
    df_print: paged
vignette: >
  %\VignetteIndexEntry{BIGpopA Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5
)
```

# Introduction

This tutorial demonstrates a typical pedigree workflow using the BIGpopA R
package, the analytical engine behind
[Familia](https://github.com/Breeding-Insight/Familia), Breeding Insight's
Shiny application for pedigree validation and ancestry assessment. We run three
functions in sequence:

* check_ped() &mdash; detect and correct structural pedigree errors
  (duplicates, conflicting trios, inconsistent sex roles, missing parents).
* validate_pedigree() &mdash; test parent&ndash;offspring trios for Mendelian
  errors against SNP genotypes and flag incorrect assignments.
* find_parentage() &mdash; assign the most likely parent(s) to progeny from a
  pool of candidate parents.

The methods are species-agnostic and apply to a wide range of plant and animal
breeding programs.

## Install

* It requires R version superior to 4.4.0

Install BIGpopA from CRAN (if not already installed):

```{r, eval=FALSE}
install.packages("BIGpopA")
```

# Input data

BIGpopA uses a small toy dataset that exercises every result category. Each
function accepts a file path or a data.frame / data.table; to keep this
vignette self-contained we build the objects inline: a pedigree, a SNP genotype
matrix coded 0/1/2, and candidate parent / progeny lists.

```{r}
library(BIGpopA)

# SNP genotypes coded 0/1/2 (NA = missing); C_low is mostly missing
genotypes = data.frame(
  id    = c("M1","F1","M2","F2","C1","C2","C3","C_bad","C_low"),
  snp01 = c(  0,   0,   2,   2,   0,   2,   1,     0,     NA),
  snp02 = c(  0,   0,   2,   0,   0,   1,   0,     0,     NA),
  snp03 = c(  0,   2,   0,   2,   1,   1,   1,     1,     NA),
  snp04 = c(  0,   2,   0,   0,   1,   0,   0,     1,     NA),
  snp05 = c(  2,   0,   1,   1,   1,   1,   1,     1,     NA),
  snp06 = c(  2,   0,   1,   0,   1,   1,   1,     1,     NA),
  snp07 = c(  2,   2,   0,   1,   2,   0,   2,     2,     NA),
  snp08 = c(  2,   2,   2,   2,   2,   2,   2,     2,     NA),
  snp09 = c(  0,   1,   0,   2,   1,   1,   1,     1,      0),
  snp10 = c(  0,   1,   2,   0,   0,   1,   0,     0,      0),
  snp11 = c(  2,   1,   1,   2,   1,   2,   2,     1,      2),
  snp12 = c(  2,   1,   0,   1,   2,   1,   1,     2,      2),
  stringsAsFactors = FALSE
)

# pedigree with founders coded 0, a mis-assigned trio (C_bad),
# and a duplicated row (C_missing)
pedigree = data.frame(
  id            = c("M1","F1","M2","F2","C1","C2","C3","C_bad","C_low","C_missing","C_missing"),
  male_parent   = c("0", "0", "0", "0", "M1","M2","M1","M2",   "M1",   "M1",       "M1"),
  female_parent = c("0", "0", "0", "0", "F1","F2","F2","F1",   "F1",   "F1",       "F1"),
  stringsAsFactors = FALSE
)

# candidate parents (with sex) and progeny to assign
parents = data.frame(
  id  = c("M1","M2","F1","F2"),
  sex = c("M", "M", "F", "F"),
  stringsAsFactors = FALSE
)

progeny = data.frame(
  id = c("C1","C2","C3","C_bad","C_low","C_missing"),
  stringsAsFactors = FALSE
)

pedigree
genotypes
```

The pedigree has four founders (M1, F1, M2, F2, both parents coded 0) and
several crosses, including a deliberately duplicated row (C_missing), a
mis-assigned trio (C_bad), and a sample with mostly missing genotypes (C_low).

# Pedigree cleaning

check_ped() operates on the pedigree columns id, male_parent, female_parent.
Exact duplicates and missing parents are always corrected; conflicting trios
and inconsistent sex roles are corrected by default. Cycles are reported only
and must be resolved manually. Set verbose = TRUE to print a full report to the
console.

```{r}
#check_ped
clean_ped_results = check_ped(pedigree, verbose = FALSE)

# the corrected, analysis-ready pedigree
clean_ped = clean_ped_results$corrected_pedigree
clean_ped
```

The returned list also exposes per-issue tables for review, e.g.
clean_ped_results$exact_duplicates (here, the repeated C_missing row) and
clean_ped_results$missing_parents.

# Pedigree validation

validate_pedigree() compares each trio against the genotype matrix and computes
a Mendelian error rate. Trios are classified as pass, fail, low_markers,
no_genotype_data, founders, or missing_parents, and a best-matching replacement
parent is suggested where relevant. With plot_results = TRUE the function draws
a histogram of trio error percentages.

```{r}
#validate_ped
ped_validate_results = validate_pedigree(clean_ped, genotypes,
                                         verbose = FALSE, plot_results = TRUE)

ped_validate_report = ped_validate_results$full_results
ped_validate_report
```

Reading the status column: C1, C2, C3 pass; C_bad fails (its recorded male
parent shows a high homozygous-mismatch rate, so the recommended correction is
to remove it); C_low is flagged low_markers because too few markers were
genotyped; and C_missing has no_genotype_data. Key tuning arguments are
trio_error_threshold (default 5.0), min_markers (default 10), and
single_parent_error_threshold (default 2.0).

# Parentage assignment

When parentage is unknown, find_parentage() searches a pool of candidate
parents for the best assignment to each progeny. The method argument selects
the search: best_pair (default), best_male_parent, best_female_parent, or
best_match.

```{r}
#find_parentage
find_parentage_results = find_parentage(genotypes, parents, progeny,
                                        method = "best_pair",
                                        verbose = FALSE, plot_results = TRUE)

find_parentage_report = find_parentage_results$full_results
find_parentage_report
```

Each progeny is assigned the pair minimizing Mendelian error and labelled pass,
high_error, or low_markers. Notice that C_bad, which failed validation against
its recorded parents, is here correctly recovered to its true pair (M1/F1) at
0% error, while C_low remains low_markers. Useful arguments include
error_threshold, min_markers, show_ties, allow_parent_selfing, and
exclude_self_match.

# Complete workflow

```{r, eval=FALSE}
clean_ped = check_ped(pedigree)$corrected_pedigree
validated = validate_pedigree(clean_ped, genotypes)$full_results
assigned  = find_parentage(genotypes, parents, progeny, method = "best_pair")$full_results
```

# Session information

```{r}
sessionInfo()
```

# How to cite

If you use BIGpopA in research, run citation("BIGpopA") for the current
reference.

```{r}
citation("BIGpopA")
```

# Acknowledgments

Developed as part of the [Breeding Insight](https://www.breedinginsight.org)
initiative to provide open-source, data-driven tools for modern breeding programs.
